Unifying quasars with clumpy wind models

### 2015-09-19 paper.tex

Approximately $20\%$ of quasars exhibit blueshifted ($\sim 0.1c$)
broad absorption lines (BALs) in the ultraviolet,
providing clear evidence for outflowing absorbing material
\citep{weymann1991, reichard2003, knigge2008, turnermiller2009, allen2011}.
The simplest explanation for the incidence of 
BAL quasars (BALQSOs) is in terms of an accretion disc wind (ADW). 
Within the ADW paradigm, a biconical wind rises from 
the accretion disc and the BALQSO fraction is associated with
the covering factor of the outflow. 
The wind is expected to be roughly equatorial from polarisation results
\citep{goodrich1995, cohen1995}.
ADWs offer a natural explanation for the
diverse phenemonology of luminous AGN and QSOs \citep[e.g.][]{MCGV95, elvis2000}. 
Depending on viewing angle, an observer 
may see a BALQSO or normal `Type 1' quasar.
Within this framework, the broad-line region (BLR) can 
correspond either to the dense wind base or dense clumps embedded
in the outflows. Indeed, \cite{elitzur2014} found that a disc-wind BLR scenario
naturally explains the emission line evolution of AGN.
A biconical wind model can also readily explain the various sub-classifications of BALQSOs: 
HiBALQSOs, which only exhibit higher ionization line absorption; LoBALQSOs which also show
absorption in lower ionization state species such as \mg\ and \al; and
FeLoBALQSOs which show further absorption in Fe~\textsc{ii} and \textsc{iii}.
In unified models, this is generally attributed to ionization stratification
of the outflow \citep[e.g.][]{elvis2000}.

@@

{\bf ksl - I a a bit torn over whether we add a few 3 sentences to describe the basic idea.  Some will be reading this for the first time.  Items that one would include are that we calculate the radiative transfer in a cylindrical coordinate system of dimension ?? x ??.  We have ionization and spectral cycles. In each cycle, ?? photon bundles are generated. We stop the ionization cycles after they converge, without any details.  I think this would be useful because we introduce terms like ``cell' later on.', and without some kind of description they have no context}

@@

Our approach to line transfer is based upon the macro-atom implementation developed by \cite{lucy2002, lucy2003}. In our case, for reasons of computational effieciency, we adopt the  hybrid macro-atom scheme 
described by M15 (an improvement over that used by H14 for our earlier quasar study), 
in which the energy flows
through the system are described in terms of indivisible
energy quanta of radiant or kinetic energy 
(`$r-$packets' and `$k-$packets' respectively; see also section 2.3).
These energy packets interact with either two-level `simple ions'
or full `macro-atoms'. 
Our scheme allows one to treat non-LTE line transfer in radiative equilibrium
without approximation for elements which are identified as 
full macro-atoms, while maintaining the fast `two-level' 
treatment of resonance lines when elements are identified 
as simple-ions (see M15). In this study,
only H \& He are treated as macro-atoms, because 
we expect recombination to be important
in determining their level populations and resultant line emission, and because we were especially interested in the contribution to AGN spectra of \LA.  In H14, all atoms were treated in a two-level approximation.

@@

%To avoid approximating the complex AGN SED as e.g. a dilute blackbody \citep[][M15]{ML93},
We model the SED in a grid cell using the technique described by H13. In this scheme,
the mean intensity in a series of $n$ bands is modeled as a normalised power law in 
frequency $\nu$
\begin{equation}
J_{\nu,i}=K_{pl}\nu^{\alpha_{pl}},
\end{equation}
for a band $i$, or an exponential 
\begin{equation}
J_{\nu,i}=K_{exp} e^{(-h\nu/k T_{exp})}.
\end{equation}
Here, $K_{exp}$, $K_{pl}$, $T_{exp}$ and $\alpha_{pl}$ are spectral fit parameters
deduced from the band-limited radiation field estimators.
The ionization rate out of ion $j$ can then be written as 
\begin{equation}
R_{j,j+1}(J)= 
\displaystyle{
n_i \left(C_{j} n_e + 
\sum_{band~i=0}^{n}~
{\int_{\nu_i}^{~\nu_i+1}{J_{\nu,i}\sigma_j(\nu)\nu^{-1}d\nu}}
\right),}
\end{equation}
where $\sigma_j$ is the photoionization cross-section and $C_{j}$
represents the collisional ionization coefficient.
The recombination rate {\em into} ion $j$ is given by 
\begin{equation}
R_{j+1,j}(T_e) = (\alpha^j_{RR} + \alpha^j_{DR} + \alpha^j_{CR}) n_{j+1} n_e,
\end{equation}
where each $\alpha^j$ here is the recombination rate coefficient into the ground state of ion $j$.
The subscripts denote radiative, dielectronic and 
collisional (three-body) recombination, respectively.
We neglect recombination to and from excited states in the simple-ion calculation. 
For simple-ions, we use a dilute Boltzmann equation to calculate 
the population of level $k$ in ionic stage $j$,
\begin{equation}
\frac{n_{jk}}{n_j} = \frac{W g_k}{z_j(T_R)} \exp(-E_k/kT_R).
\end{equation}
Here $z_j$ is the partition function of ionic stage $j$,
$T_R$ is the effective radiation temperature,
$E_k$ is the energy difference between level $k$ and the ground state,
and $g_k$ is the statistical weight of level $k$.

@@

@@

\begin{figure*} %fullpage
\centering
\includegraphics[width=1.0\textwidth]{figures/uvspec.png}
\caption
{
Synthetic spectra at five viewing angles in our. The coloured lines
show different quasar and BAL quasar composites, and the dotted line shows a disc
only continuum to show the effect of the outflow on the continuum level.  
}
\label{fig:uvspec}
\end{figure*} %fullpage

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{figures/windnew3.png}
\caption
{
A cartoon showing the broad classes of sightlines 
in our model, illustrates how geometric effects lead to 
the different emergent spectra. The colour gradient is approximate,
but indicates the stratified ionization structure, 
from highly ionized (yellow) to low ionization (purple) material.
%[FIGURE NEEDS IMPROVING]
}
\label{fig:sightline}
\end{figure}

@@

Figure~\ref{fig:uvspec} shows the synthetic spectrum in the UV from our model. 
We also show a comparison to composite quasar and BALQSO spectra.
We show a cartoon illustrating how geometric effects determine
the output spectra in figure~\ref{fig:sightline}.  {\bf - Need to explain the source of the BAL composites}

@@

We find that the model can produce significant line emission
at low inclinations, particular in \civ, 
and the improved treatment of recombination results in a strong \la\
line. In the context of unification, this is a promising result, 
and shows that a biconical wind can produce significant 
emission at `quasar-like' angles. To assess the ability of the model to match real 
quasar spectra, we also show {\sl Sloan Digital Sky Survey} (SDSS) quasar composites from
\cite{reichard2003}, normalised to the flux at 2000\AA\ in each panel. Our simulated spectra do not contain
the semi-forbidden 
intercombination lines seen in quasar spectra because our radiative transfer code 
currently does not have a treatment for semi-forbidden lines.  {\bf Isn't this just a statement about the fact that we have not implemented CIII as a full macro-atom; if so we should say that}
This is especially noticeable with the strong C~\textsc{iii}]~1909\AA\, 
line in the quasar composite spectra.  
The critical density of the C~\textsc{iii}]~1909\AA\, line 
is $n_e\sim10^{9.5}~\rm{cm^{-3}}$, which is higher than much of the 
outer portion of our wind. We therefore expect intercombination lines
to become important coolants in the outer portion of the wind. {\bf we have gone from their being lines we should see, to important coolants.  I think we should stick to the fact that we would think our code would see them if we put C in as a macro atom.  Also I think this who discussion of CIII should be after we have discussed what looks right with the emission lines.}

@@

When comparing BALQSO and quasar composites, it is apparent
that they possess remarkably similar line strengths and widths 
\citep[e.g.][]{reichard2003}.  {\bf I think we mean EW here; otherwise I don't understand what we are talking about.  We could say ``line shapes and equivalent widths''.  I also think this requries more than one sentence of introduction.}
This presents a challenge to our model, as well as the geometric 
unification picture in general, because these quantities should vary with inclination angle.
Limb darkening and foreshortening causes the disc continuum emission to be 
strongly anisotropic, yet the line emission in our models is much more isotropic. 
This has the effect of enhancing the line-to-continuum ratios at high inclinations. 
To construct a scenario where emission 
line equivalent widths are comparable at all inclinations requires 
significant fine-tuning when considering a classical thin disc. {\bf Is the previous sentence a throwaway line, or has some person done this.  If so, please reference; if not please delete.}
%An alternative, simpler solution is that the emergent continuum is roughly isotropic.

Obviously,  one possible solution to this problem is hypothesize a more isotropic distribution for the emergent condition than predicted by a ``classical'' disc. 
General relativistic effects -- specifically, light bending
and relativistic beaming -- can cause 
the accretion disc SED to become more isotropic \citep[e.g.][]{zhang1997,munozdarias2013}.  Incorporating these effects into {\sc PYTHON} would be a very difficult and time-consuming challenge. Therefore, for the purposes of this discussion we have chosen to investigate whether GR effects could produce enough isotropy to address the EW problem, we have generated 
\ GR disc spectra, we use the code \agn\ \citep{hubeny2000,davishubeny2006,davis2007}. 
The output flux at three different wavelengths
as a function of inclination for an \agn\ model with the same disc and BH parameters
as our clumpy wind model is shown in figure~\ref{fig:f2000}.  The results indicate that 
the effects of GR on an AGN disc are much less extreme 
in the UV portion of the spectrum than the calculations
by \cite{zhang1997} for the X-rays in X-ray binaries.
The emission for the disks of AGN is still strongly anisotropic.
GR alone therefore cannot explain the line ratio trends in quasars by making the disc
emit more isotropically.
This supports the findings of \cite{risaliti2011}, who find that EW distributions 
in quasars are consistent with anisotropic emission from optically thick, 
disc-like structures for {\em both} the continuum source and BLR. 
We will explore these ideas further in a future study.  {\bf Wait!  All this says is that there has to be an alternative explanation.  It is not support for any specific explanation!. This last few sentences should be a separate paragraph.  They should start out: An alernative explanation proposed by risalite is that...}

>>>>

Approximately $20\%$ of quasars exhibit blueshifted ($\sim 0.1c$)
broad absorption lines (BALs) in the ultraviolet,
providing clear evidence for outflowing absorbing material
\citep{weymann1991, reichard2003, knigge2008, turnermiller2009, allen2011}.
The simplest explanation for the incidence of 
BAL quasars (BALQSOs) is in terms of an accretion disc wind (ADW). 
Within the ADW paradigm, a biconical wind rises from 
the accretion disc and the BALQSO fraction is associated with
the covering factor of the outflow. 
Polarisation studies expect the wind to be roughly equatorial
\citep{goodrich1995, cohen1995}.
ADWs offer a natural explanation for the
diverse phenemonology of luminous AGN and QSOs \citep[e.g.][]{MCGV95, elvis2000}. 
Depending on viewing angle, an observer 
may see a BALQSO or normal `Type 1' quasar.
Within this framework, the broad-line region (BLR) can 
correspond either to the dense wind base or dense clumps embedded
in the outflows. Indeed, \cite{elitzur2014} found that a disc-wind BLR scenario
naturally explains the emission line evolution of AGN.
A biconical wind model can also readily explain the various sub-classifications of BALQSOs: 
HiBALQSOs, which only exhibit high ionization line absorption; LoBALQSOs, which also show
absorption in lower ionization state species such as \mg\ and \al; and
FeLoBALQSOs which show further absorption in Fe~\textsc{ii} and \textsc{iii}.
In unified models, this is generally attributed to ionization stratification
of the outflow \citep[e.g.][]{elvis2000}.

@@

{\bf ksl - I a a bit torn over whether we add a few 3 sentences to describe the basic idea.  Some will be reading this for the first time.  Items that one would include are that we calculate the radiative transfer in a cylindrical coordinate system of dimension ?? x ??.  We have ionization and spectral cycles. In each cycle, ?? photon bundles are generated. We stop the ionization cycles after they converge, without any details.  I think this would be useful because we introduce terms like ``cell'' later on.', and without some kind of description they have no context}

@@

Our approach to line transfer is based upon the macro-atom implementation developed by \cite{lucy2002, lucy2003}. In our case, for reasons of computational effieciency, we adopt the  hybrid macro-atom scheme 
described by M15 (an improvement over that used by H14 for our earlier quasar study), 
in which the energy flows
through the system are described in terms of indivisible
energy quanta of radiant or kinetic energy 
(`$r-$packets' and `$k-$packets' respectively; see also section 2.3).
These energy packets interact with either two-level `simple ions'
or full `macro-atoms'. 
Our scheme allows one to treat non-LTE line transfer in radiative equilibrium
without approximation for elements which are identified as 
full macro-atoms, while maintaining the fast `two-level' 
treatment of resonance lines when elements are identified 
as simple-ions (see M15). In this study,
only H is treated as a macro-atom, because 
we expect recombination to be important
in determining their level populations and resultant line emission, and because we were especially interested in the contribution to AGN spectra of \LA.  In H14, all atoms were treated in a two-level approximation.

@@

%To avoid approximating the complex AGN SED as e.g. a dilute blackbody \citep[][M15]{ML93},
We model the SED in a grid cell using the technique described by H13. In this scheme,
the mean intensity in a series of $n$ bands is modeled as a normalised power law in 
frequency $\nu$
\begin{equation}
J_{\nu,i}=K_{pl}\nu^{\alpha_{pl}},
\end{equation}
for a band $i$, or an exponential 
\begin{equation}
J_{\nu,i}=K_{exp} e^{(-h\nu/k T_{exp})}.
\end{equation}
Here, $K_{exp}$, $K_{pl}$, $T_{exp}$ and $\alpha_{pl}$ are spectral fit parameters
deduced from the band-limited radiation field estimators.
The ionization rate out of ion $j$ can then be written as 
\begin{equation}
R_{j,j+1}(J)= 
\displaystyle{
n_i \left(C_{j} n_e + 
\sum_{band~i=0}^{n}~
{\int_{\nu_i}^{~\nu_i+1} \frac{4 \pi J_{\nu,i}\sigma_j(\nu)}  {h \nu} d\nu}
\right),}
\end{equation}
where $\sigma_j$ is the photoionization cross-section and $C_{j}$
represents the collisional ionization coefficient.
The recombination rate {\em into} ion $j$ is given by 
\begin{equation}
R_{j+1,j}(T_e) = (\alpha^j_{RR} + \alpha^j_{DR} + \alpha^j_{CR}) n_{j+1} n_e,
\end{equation}
where each $\alpha^j$ here is the recombination rate coefficient into the ground state of ion $j$.
The subscripts denote radiative, dielectronic and 
collisional (three-body) recombination, respectively.
%% We neglect recombination to and from excited states in the simple-ion calculation. 
For simple-ions, we use a dilute Boltzmann equation to calculate 
the population of level $k$ in ionic stage $j$,
\begin{equation}
\frac{n_{jk}}{n_j} = \frac{W g_k}{z_j(T_R)} \exp(-E_k/kT_R).
\end{equation}
Here $z_j$ is the partition function of ionic stage $j$,
$T_R$ is the effective radiation temperature,
$E_k$ is the energy difference between level $k$ and the ground state,
and $g_k$ is the statistical weight of level $k$.

@@

Our previous modelling efforts assumed a smooth outflow, 
in which the density at a given point was determined only by the 
kinematic parameters and mass loss rate. However, as already discussed,
AGN winds exhibit significant substructure -- the outflow is expected to be
{\em clumpy}, rather than smooth, and probably on a variety of scales. 
Implementing a treatment of clumping is challenging, for
two main reasons. First, there are significant 
computational difficulties associated with adequately
resolving and realistically modelling a series of small scale, high density
regions with a MCRT code. Second, the addition of multiple additional degrees of
freedom in the model results in significantly wider parameter space.
Unfortuantely, the physical scale lengths and density contrasts 
associated with these parameters cannot be well-constrained from observations.

@@

\begin{figure*} %fullpage
\centering
\includegraphics[width=1.0\textwidth]{figures/uvspec.png}
\caption
{
Synthetic spectra at five viewing angles in our model. The coloured lines
show different quasar and BAL quasar composites, and the dotted line shows a disc
only continuum to show the effect of the outflow on the continuum level.  
}
\label{fig:uvspec}
\end{figure*} %fullpage

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{figures/windnew3.png}
\caption
{
A cartoon describing the broad classes of sightline 
in our model, illustrating how geometric effects lead to 
the different emergent spectra. The colour gradient is approximate,
but indicates the stratified ionization structure, 
from highly ionized (yellow) to low ionization (purple) material.
%[FIGURE NEEDS IMPROVING]
}
\label{fig:sightline}
\end{figure}

@@

Figure~\ref{fig:uvspec} shows the synthetic spectrum in the UV from our model. 
To assess the ability of the model to match real 
quasar spectra, we also show {\sl Sloan Digital Sky Survey} (SDSS) quasar and BALQSO 
composites from \cite{reichard2003}, normalised to the flux at 2000\AA\ in each panel. 
We show a cartoon illustrating how geometric effects determine
the output spectra in figure~\ref{fig:sightline}.

@@

We find that the model can produce significant line emission
at low inclinations, particular in \civ, 
and the improved treatment of recombination results in a strong \la\
line. In the context of unification, this is a promising result, 
and shows that a biconical wind can produce significant 
emission at `quasar-like' angles. 
Our simulated spectra do not contain the strong C~\textsc{iii}]~1909\AA\, 
line seen in the quasar composite spectra. This is because we do not yet
treat C as a full macro-atom with a full collisional rates between forbidden
or semi-forbidden transitions, as would be required.
The critical density of the C~\textsc{iii}]~1909\AA\, line 
is $n_e\sim10^{9.5}~\rm{cm^{-3}}$, which is higher than much of the 
outer portion of our wind. We therefore expect a model with these 
parameters to produce C~\textsc{iii}]~1909\AA\ with a proper treament. 
% {\bf we have gone from their being lines we should see, to important coolants.  I think we should stick to the fact that we would think our code would see them if we put C in as a macro atom.  Also I think this who discussion of CIII should be after we have discussed what looks right with the emission lines.}

@@

When comparing BALQSO and quasar composites, it is apparent
that they possess remarkably similar line strengths and widths 
\citep[e.g.][]{weymann1991,reichard2003}.  
{\bf I think we mean EW here; otherwise I don't understand what we are talking about.  We could say ``line shapes and equivalent widths''.  I also think this requries more than one sentence of introduction.}
>>>>>>> 3c3265bb9688e4788107378165d92e06c2aa7abe
This presents a challenge to our model, as well as the geometric 
unification picture in general, because these quantities should vary with inclination angle.
Limb darkening and foreshortening causes the disc continuum emission to be 
strongly anisotropic, yet the line emission in our models is much more isotropic. 
This has the effect of enhancing the line-to-continuum ratios at high inclinations. 
To construct a scenario where emission 
line equivalent widths are comparable at all inclinations requires 
significant fine-tuning when considering a classical thin disc. {\bf Is the previous sentence a throwaway line, or has some person done this.  If so, please reference; if not please delete.}
%An alternative, simpler solution is that the emergent continuum is roughly isotropic.

Obviously,  one possible solution to this problem is hypothesize a more isotropic distribution for the emergent condition than predicted by a ``classical'' disc. 
General relativistic effects -- specifically, light bending
and relativistic beaming -- can cause 
the accretion disc SED to become more isotropic \citep[e.g.][]{zhang1997,munozdarias2013}.  Incorporating these effects into {\sc PYTHON} would be a very difficult and time-consuming challenge. Therefore, for the purposes of this discussion we have chosen to investigate whether GR effects could produce enough isotropy to address the EW problem, we have generated 
\ GR disc spectra, we use the code \agn\ \citep{hubeny2000,davishubeny2006,davis2007}. 
The output flux at three different wavelengths
as a function of inclination for an \agn\ model with the same disc and BH parameters
as our clumpy wind model is shown in figure~\ref{fig:f2000}.  The results indicate that 
the effects of GR on an AGN disc are much less extreme 
in the UV portion of the spectrum than the calculations
by \cite{zhang1997} for the X-rays in X-ray binaries.
The emission for the disks of AGN is still strongly anisotropic.
GR alone therefore cannot explain the line ratio trends in quasars by making the disc
emit more isotropically. This supports the findings of \cite{risaliti2011}, 
who find that EW distributions in quasars are consistent with anisotropic emission 
from optically thick, 
disc-like structures for {\em both} the continuum source and BLR. 
If this is the case, it has a dramatic affect on the intrinsic
BAL fraction inferred from flux-limited samples \citep{goodrich1997,krolikvoit1998}.
We will explore these ideas further in a future study.
We will explore these ideas further in a future study.  {\bf Wait!  All this says is that there has to be an alternative explanation.  It is not support for any specific explanation!. This last few sentences should be a separate paragraph.  They should start out: An alernative explanation proposed by risalite is that...}
